---
title: "Statistical analyses of time-series environmental correlates and selected taxa"
knit: (function(input_file, encoding) {
  out_dir <- '.';
  rmarkdown::render(input_file,
  encoding=encoding,
  output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "rhiannon mondav"
date: "03/12/2019"
output: html_document
---

```{r setup, include=FALSE}
r <- getOption("repos")
r["CRAN"] <-"https://mirrors.dotsrc.org/cran/"
options(repos = r)
## install required packages
if(!require(beanplot)){
  install.packages("beanplot")
}

if(!require(pgirmess)){
  devtools::install_github("pgirmess")
}

if(!require(usdm)){
  devtools::install_github("usdm")
}

if(!require(scales)){
  devtools::install_github("scales")
}

if(!require(psych)){
  devtools::install_github("psych")
}

## load required packages used in first section
library(usdm)
library(scales)
library(beanplot)
library(pgirmess)

knitr::opts_chunk$set(echo = TRUE)
```



```{r}
## read in metadata table
TS_mapALL <-
  as.data.frame(read.delim("./Mondav_etal_2020_Erken_network/data/timeseries_map_allwaterdata.txt",row.names=1))

## check metadata for colinearity
## first check basics
summary(TS_mapALL)
#remove columns 29-33 as high NAs
TS_mapRDCD<-(TS_mapALL[,-c(29:33)])
## test columns 8 to 28 as pre 8 is seq data
## set maximum correlation at 10
vifstep_env<-vifstep(TS_mapRDCD[,c(10:28)],th=10)
vifstep_env
#remove columns 27,28 as colinear
TS_mapRDCD<-(TS_mapRDCD[,-c(27:28)])
## now re-test with th emore sensitive vifcor
## set maximum correlation at 0.7
vifcor_env<-vifcor(TS_mapRDCD[,c(10:26)],th=0.7)
vifcor_env
summary(TS_mapRDCD)
## remove colinear variables identified 
TS_mapRDCD<-(TS_mapRDCD[,-c(23:25)])
vifcor_env<-vifcor(TS_mapRDCD[,c(10:23)],th=0.7)
vifcor_env

```


## Figure S8
## plot factors etc for publication and sanity check
## beanplots of the remaining 14 environmental parameters verses lake layer

```{r beanplots, echo=FALSE}
## assign plotting shape from map file
seqshape=as.character(TS_mapRDCD$seq_shape)
## assign the main factors we will be testing
TS_mapRDCD$Layer<-
  factor(TS_mapRDCD$Layer, levels=c('epi','meta','hypo','mix-C','mix-I'))
timeseries_season_order<-
  ordered(TS_mapRDCD$season,
          levels=c('winter','spring','epi','meta','hypo','autumn'))

## alpha is a variable in both psych and scales
## unload psych just in case already loaded
detach("package:psych", unload=TRUE)
library(scales)
par(mar=c(4.5,4.5,0.5,0.5),oma=c(0,0,0,0),mfrow=c(5,3),las=2)
beanplot(TS_mapRDCD$secchi_m~TS_mapRDCD$season, ylab='log secchi(m)',
         what=c(1,1,1,0),col=alpha("grey30",0.5),border =NA,cex.lab=1.2)
beanplot(TS_mapRDCD$pH~TS_mapRDCD$season, ylab='log pH',
         what=c(1,1,1,0),col=alpha("grey30",0.5),border =NA,cex.lab=1.2)
beanplot(TS_mapRDCD$Alk_mekw_per_L~TS_mapRDCD$season, ylab='log Alk (mekw/L)',
         what=c(1,1,1,0),col=alpha("grey30",0.5),border =NA,cex.lab=1.2)

beanplot(TS_mapRDCD$Cond_mS_per_m~TS_mapRDCD$season, ylab='cond (mS/L)',
         what=c(1,1,1,0),col=alpha("grey30",0.5),border =NA,cex.lab=1.2)
beanplot(TS_mapRDCD$Col_mgPt_per_L~TS_mapRDCD$season, ylab='log col (mgPt/L)',
         what=c(1,1,1,0),col=alpha("grey30",0.5),border =NA,cex.lab=1.2)
beanplot(TS_mapRDCD$turbidity_FNU_EV26~TS_mapRDCD$season, ylab='log turb',
         what=c(1,1,1,0),col=alpha("grey30",0.5),border =NA,cex.lab=1.2)

beanplot(TS_mapRDCD$TotP_ug_per_L~TS_mapRDCD$season, ylab='log TotP (ug/L)',
         what=c(1,1,1,0),col=alpha("grey30",0.5),border =NA,cex.lab=1.2)
beanplot(TS_mapRDCD$NH4_ug_per_L~TS_mapRDCD$season, ylab='NH4 (ug/L)',
         what=c(1,1,1,0),col=alpha("grey30",0.5),border =NA,cex.lab=1.2)
beanplot(TS_mapRDCD$NO2andNO3_ug_per_L~TS_mapRDCD$season, ylab='NO2+NO3 (ug/L)',
         what=c(1,1,1,0),col=alpha("grey30",0.5),border =NA,cex.lab=1.2)

beanplot(TS_mapRDCD$PN_ug_per_L~TS_mapRDCD$season, ylab='log PN (ug/L)',
         what=c(1,1,1,0),col=alpha("grey30",0.5),border =NA,cex.lab=1.2)
beanplot(TS_mapRDCD$TotN_ug_per_L~TS_mapRDCD$season, ylab='log TotN (ug/L)',
         what=c(1,1,1,0),col=alpha("grey30",0.5),border =NA,cex.lab=1.2)
beanplot(TS_mapRDCD$Si_ug_per_L~TS_mapRDCD$season, ylab='Si (ug/L)',
         what=c(1,1,1,0),col=alpha("grey30",0.5),border =NA,cex.lab=1.2)

beanplot(TS_mapRDCD$Temp_C~TS_mapRDCD$season, ylab='Temp (°C)',
         what=c(1,1,1,0),col=alpha("grey30",0.5),border =NA,cex.lab=1.2)
beanplot(TS_mapRDCD$O2pcnt~TS_mapRDCD$season, ylab='O2 (mg/L)',
         what=c(1,1,1,0),col=alpha("grey30",0.5),border =NA,cex.lab=1.2)
```


## now an actual stats test 
## to check for significant difference of association between layers
```{r}
kruskal.test(TS_mapRDCD$av_depth~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$secchi_m~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$pH~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$Alk_mekw_per_L~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$Cond_mS_per_m~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$Col_mgPt_per_L~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$turbidity_FNU_EV26~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$NH4_ug_per_L~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$NO2andNO3_ug_per_L~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$PN_ug_per_L~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$TotN_ug_per_L~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$Si_ug_per_L~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$Temp_C~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$O2pcnt~TS_mapRDCD$season)
## all except alkalinity had p values <0.001 diff with season
## alkalinity had prob of diff of 0.006 ie p<0.01  
## so test all with post-hoc tests to ID which env parameters are sig.diff  
## identify those with at least two with sig diferences
kruskalmc(TS_mapRDCD$secchi_m~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$pH~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$Alk_mekw_per_L~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$Cond_mS_per_m~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$Col_mgPt_per_L~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$turbidity_FNU_EV26~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$TotP_ug_per_L~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$NH4_ug_per_L~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$NO2andNO3_ug_per_L~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$PN_ug_per_L~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$TotN_ug_per_L~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$Si_ug_per_L~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$Temp_C~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$O2pcnt~TS_mapRDCD$season,probs=0.001)
```

## this section looks at taxa abundances and seasonal differences
## load up output files from the summarize taxa script by QIIME
```{r}
## uses the output from 'summarize taxa' in QIIME
## files are in species (rows) by sample (columns) format
TS_N1500_genera <-read.delim("./Mondav_etal_2020_Erken_network/data/otu_table_N1500_L6.txt", row.names=1)
TS_N1500_family <-read.delim("./Mondav_etal_2020_Erken_network/data/otu_table_N1500_L5.txt", row.names=1)
TS_N1500_order <-read.delim("./Mondav_etal_2020_Erken_network/data/otu_table_N1500_L4.txt", row.names=1)
TS_N1500_class <-read.delim("./Mondav_etal_2020_Erken_network/data/otu_table_N1500_L3.txt", row.names=1)
TS_N1500_phyla <-read.delim("./Mondav_etal_2020_Erken_network/data/otu_table_N1500_L2.txt", row.names=1)
## transform the tables to be sample by species format
generat<-as.data.frame(t(TS_N1500_genera))
familyt<-as.data.frame(t(TS_N1500_family))
ordert<-as.data.frame(t(TS_N1500_order))
classt<-as.data.frame(t(TS_N1500_class))
phylat<-as.data.frame(t(TS_N1500_phyla))

## input abundances of taxa of interest as columns in metadata file
TS_mapRDCD$LD12<-100*generat$`k__Bacteria;p__Proteobacteria;c__Alphaproteobacteria;o__Rickettsiales;f__Pelagibacteraceae;g__LD12`
TS_mapRDCD$acI<-100*familyt$`k__Bacteria;p__Actinobacteria;c__Actinobacteria;o__Actinomycetales;f__ACK-M1`
TS_mapRDCD$C111<-100*familyt$`k__Bacteria;p__Actinobacteria;c__Acidimicrobiia;o__Acidimicrobiales;f__C111`
TS_mapRDCD$Comamonadaceae<-100*familyt$`k__Bacteria;p__Proteobacteria;c__Betaproteobacteria;o__Burkholderiales;f__Comamonadaceae`
TS_mapRDCD$limnohab<-100*generat$`k__Bacteria;p__Proteobacteria;c__Betaproteobacteria;o__Burkholderiales;f__Comamonadaceae;g__Limnohabitans`
TS_mapRDCD$flavo<-100*generat$`k__Bacteria;p__Bacteroidetes;c__Flavobacteriia;o__Flavobacteriales;f__Flavobacteriaceae;g__Flavobacterium`
TS_mapRDCD$Verrucomicrobia<-100*phylat$`k__Bacteria;p__Verrucomicrobia`
TS_mapRDCD$Chloroplast<-100*classt$`k__Bacteria;p__Cyanobacteria;c__Chloroplast`

## phytos
TS_mapRDCD$synechococcus<-100*generat$`k__Bacteria;p__Cyanobacteria;c__Synechococcophycideae;o__Synechococcales;f__Synechococcaceae;g__Synechococcus`
TS_mapRDCD$nostocaceae<-100*familyt$`k__Bacteria;p__Cyanobacteria;c__Nostocophycideae;o__Nostocales;f__Nostocaceae`
TS_mapRDCD$microcystis<-100*generat$`k__Bacteria;p__Cyanobacteria;c__Oscillatoriophycideae;o__Chroococcales;f__Microcystaceae;g__Microcystis`
TS_mapRDCD$haptophyceae<-100*familyt$`k__Bacteria;p__Cyanobacteria;c__Chloroplast;o__Haptophyceae;f__`
TS_mapRDCD$cryptophyta<-100*familyt$`k__Bacteria;p__Cyanobacteria;c__Chloroplast;o__Cryptophyta;f__`
TS_mapRDCD$streptphyt<-100*generat$`k__Bacteria;p__Cyanobacteria;c__Chloroplast;o__Streptophyta;f__;g__`
TS_mapRDCD$stramenopiles<-100*ordert$`k__Bacteria;p__Cyanobacteria;c__Chloroplast;o__Stramenopiles`
TS_mapRDCD$phytos<-100*phylat$`k__Bacteria;p__Cyanobacteria`
#write.table(TS_mapRDCD,col.names = T,file = "./Mondav_etal_2020_Erken_network/data/TS_mapRDCD.txt",sep ='\t',quote=FALSE)
#TS_mapRDCD <- as.data.frame(read.delim("./Mondav_etal_2020_Erken_network/data/TS_mapRDCD.txt", row.names=1))

## test for correlation of taxa and measured environ parameters (not colinear)
library(psych)
## inteference between varibles so unload scales before using psych package
detach("package:scales", unload=TRUE)
cor_LD12env_taxa<-corr.test(TS_mapRDCD$LD12,TS_mapRDCD[,c(9:39)],
                            method = "pearson",adjust = "BH",alpha = 0.001,ci=F)
cor_acIenv_taxa<-corr.test(TS_mapRDCD$acI,TS_mapRDCD[,c(9:39)],
                       method = "pearson",adjust = "BH",alpha = 0.001,ci=F)
cor_C111env_taxa<-corr.test(TS_mapRDCD$C111,TS_mapRDCD[,c(9:39)],
                        method = "pearson",adjust = "BH",alpha = 0.001,ci=F)
cor_stramenenv_taxa<-corr.test(TS_mapRDCD$stramenopiles,TS_mapRDCD[,c(9:39)],
                        method = "pearson",adjust = "BH",alpha = 0.001,ci=F)
cor_comamenv_taxa<-corr.test(TS_mapRDCD$Comamonadaceae,TS_mapRDCD[,c(9:39)],
            method = "pearson",adjust = "BH",alpha = 0.001,ci=F)
cor_limnohabenv_taxa<-corr.test(TS_mapRDCD$limnohab,TS_mapRDCD[,c(9:39)],
            method = "pearson",adjust = "BH",alpha = 0.001,ci=F)
cor_flavoenv_taxa<-corr.test(TS_mapRDCD$flavo,TS_mapRDCD[,c(9:39)],
            method = "pearson",adjust = "BH",alpha = 0.001,ci=F)
cor_phytosenv_taxa<-corr.test(TS_mapRDCD$phytos,TS_mapRDCD[,c(9:39)],
            method = "pearson",adjust = "BH",alpha = 0.001,ci=F)


write.table(cor_LD12env_taxa$r,col.names = T,quote=FALSE,sep ='\t',
            file = "./Mondav_etal_2020_Erken_network/data/LD12env_taxaPcor.txt")
write.table(cor_LD12env_taxa$p,col.names = T,quote=FALSE,sep ='\t',
            file = "./Mondav_etal_2020_Erken_network/data/LD12env_taxaPpval.txt")
write.table(cor_acIenv_taxa$r,col.names = T,sep ='\t',quote=FALSE
            ,file = "./Mondav_etal_2020_Erken_network/data/acIenv_taxaPcor.txt")
write.table(cor_acIenv_taxa$p,col.names = T,sep ='\t',quote=FALSE
            ,file = "./Mondav_etal_2020_Erken_network/data/acIenv_taxaPpval.txt")
write.table(cor_C111env_taxa$r,col.names = T,sep ='\t',quote=FALSE
            ,file = "./Mondav_etal_2020_Erken_network/data/C111env_taxaPcor.txt")
write.table(cor_C111env_taxa$p,col.names = T,sep ='\t',quote=FALSE
            ,file = "./Mondav_etal_2020_Erken_network/data/C111env_taxaPpval.txt")
write.table(cor_stramenenv_taxa$r,col.names = T,sep ='\t',quote=FALSE
            ,file = "./Mondav_etal_2020_Erken_network/data/stramenenv_taxaPcor.txt")
write.table(cor_stramenenv_taxa$p,col.names = T,sep ='\t',quote=FALSE
            ,file = "./Mondav_etal_2020_Erken_network/data/stramenenv_taxaPpval.txt")
write.table(cor_comamenv_taxa$r,col.names = T,sep ='\t',quote=FALSE
            ,file = "./Mondav_etal_2020_Erken_network/data/comamenv_taxaPcor.txt")
write.table(cor_comamenv_taxa$p,col.names = T,sep ='\t',quote=FALSE
            ,file = "./Mondav_etal_2020_Erken_network/data/comamenv_taxaPpval.txt")
write.table(cor_limnohabenv_taxa$r,col.names = T,sep ='\t',quote=FALSE
            ,file = "./Mondav_etal_2020_Erken_network/data/limnohabenv_taxaPcor.txt")
write.table(cor_limnohabenv_taxa$p,col.names = T,sep ='\t',quote=FALSE
            ,file = "./Mondav_etal_2020_Erken_network/data/limnohabenv_taxaPpval.txt")
write.table(cor_flavoenv_taxa$r,col.names = T,quote=FALSE,
            file = "./Mondav_etal_2020_Erken_network/data/flavoenv_taxaPcor.txt",sep ='\t')
write.table(cor_flavoenv_taxa$p,col.names = T,quote=FALSE,
            file = "./Mondav_etal_2020_Erken_network/data/flavoenv_taxaPpval.txt",sep ='\t')
write.table(cor_phytosenv_taxa$r,col.names = T,quote=FALSE,
            file = "./Mondav_etal_2020_Erken_network/data/phytosenv_taxaPcor.txt",sep ='\t')
write.table(cor_phytosenv_taxa$p,col.names = T,quote=FALSE,
            file = "./Mondav_etal_2020_Erken_network/data/phytosenv_taxaPpval.txt",sep ='\t')
```


## Figure S3 scatter plots
```{r taxa env plots, echo=FALSE}
## and now some scatter plots of the env correlates calculated above
## first calculate the correlation line
acI_depth_line<-(lm(TS_mapRDCD$av_depth~TS_mapRDCD$acI))
acI_O2_line<-(lm(TS_mapRDCD$O2pcnt~TS_mapRDCD$acI))
LD12_NOx_line<-(lm(TS_mapRDCD$NO2andNO3_ug_per_L~TS_mapRDCD$LD12))
LD12_Temp_line<-(lm(TS_mapRDCD$Temp_C~TS_mapRDCD$LD12))
## set bottom and left graph margins large for labels m(b,l,t,r)
## set outer plot margins low
par(mar=c(4,4,1,1),oma=c(0,0,0,0),mfrow=c(2,2))
plot(TS_mapRDCD$av_depth,TS_mapRDCD$acI,
     xlab="av. depth (m)",ylab="acI %r.a.",
     cex.lab=1.3,pch=TS_mapRDCD$seq_shape)
abline(acI_depth_line,col="blue",lwd=1.5)

plot(TS_mapRDCD$O2pcnt,TS_mapRDCD$acI,
     xlab="O2 (%)",ylab="acI %r.a.",
     cex.lab=1.3,pch=TS_mapRDCD$seq_shape)
abline(acI_O2_line,col="blue",lwd=1.5)

plot(TS_mapRDCD$NO2andNO3_ug_per_L,TS_mapRDCD$LD12,
     xlab="NO2+NO3 (ug/L)",ylab="LD12 %r.a.",col="grey60",
     cex.lab=1.3,pch=TS_mapRDCD$seq_shape)
abline(LD12_NOx_line,col="blue",lwd=1.5)

plot(TS_mapRDCD$Temp_C,TS_mapRDCD$LD12,
     xlab="Temp (°C)",ylab="LD12 %r.a.", 
     cex.lab=1.3,pch=TS_mapRDCD$seq_shape)
abline(LD12_Temp_line,col="blue",lwd=1.5)
```


## Figure S2
```{r taxa season beanplots, echo=FALSE}
## this section deals with taxa abundance and lake cycle
detach("package:psych", unload=TRUE)
library(scales)
## get taxon names from IDs with vlookup in spreadsheet and upload
#taxa_corr<-read.delim("./Mondav_etal_2020_Erken_network/data/taxa_corr.txt", row.names=1)

## beanplots of taxa versus cycle
par(mar=c(4,4,2,2),oma=c(1,1,0,0),mfrow=c(3,2))
beanplot(TS_mapRDCD$acI~TS_mapRDCD$season,
         ylab='acI % rel abund',what=c(1,1,1,0), 
         col=alpha("grey30",0.5), border=NA,log="")
beanplot(TS_mapRDCD$stramenopiles~TS_mapRDCD$season,
         ylab='Stramenopiles % rel abund',what=c(1,1,1,0), 
         col=alpha("grey30",0.5), border=NA,bw="nrd0")
beanplot(TS_mapRDCD$C111~TS_mapRDCD$season,
         ylab='acIV/C111 % rel abund',what=c(1,1,1,0), 
         col=alpha("grey30",0.5), border=NA,log="")
beanplot(TS_mapRDCD$Comamonadaceae~TS_mapRDCD$season,
         ylab='Comamonadaceae % rel abund',what=c(1,1,1,0), 
         col=alpha("grey30",0.5), border=NA,log="")
beanplot(TS_mapRDCD$LD12~TS_mapRDCD$season,
         ylab='Ca.Fonsibacter % rel abund',what=c(1,1,1,0), 
         col=alpha("grey30",0.5), border=NA,log="")
beanplot(TS_mapRDCD$phytos~TS_mapRDCD$season,
         ylab='photosynths % rel abund',what=c(1,1,1,0), 
         col=alpha("grey30",0.5), border=NA,bw="nrd0",log="")
```


```{r}
## and do the statistical tests for the plots
kruskal.test(TS_mapRDCD$acI~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$stramenopiles~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$C111~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$Comamonadaceae~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$LD12~TS_mapRDCD$season)
kruskal.test(TS_mapRDCD$phytos~TS_mapRDCD$season)
kruskalmc(TS_mapRDCD$acI~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$stramenopiles~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$C111~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$Comamonadaceae~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$LD12~TS_mapRDCD$season,probs=0.001)
kruskalmc(TS_mapRDCD$phytos~TS_mapRDCD$season,probs=0.001)
```

```{r}
savehistory("./Mondav_etal_2020_Erken_network/data/env.Rhistory")
save.image("./Mondav_etal_2020_Erken_network/data/erken_network.RData")
```

